---
title: "README"
author: "Lukas DiGiovanni"
date: "2025-02-26"
output: html_document
---

## Uploading the Dataset

```{r}
# Install required packages if not already installed
if (!require(curl)) install.packages("curl", dependencies = TRUE)
if (!require(utils)) install.packages("utils", dependencies = TRUE)

# Load the packages
library(curl)
library(utils)

# Define the URL from Kaggle
dataset_url <- "https://www.kaggle.com/api/v1/datasets/download/fratzcan/usa-house-prices"

# Set the output file paths
zip_file <- "./usa-house-prices.zip"
extract_dir <- "./usa-house-prices"

# Download the ZIP file
curl_download(dataset_url, zip_file)

# Create extraction directory if it doesn't exist
if (!dir.exists(extract_dir)) dir.create(extract_dir)

# Unzip the file
unzip(zip_file, exdir = extract_dir)

# List extracted files
extracted_files <- list.files(extract_dir, full.names = TRUE)
message("Download and extraction complete! Files saved at: ", extract_dir)
message("Extracted files: ", paste(extracted_files, collapse = ", "))

```

```{r}
setwd("/Users/lukasdigiovanni/Desktop/DS_201")

data <- read.csv("USA Housing Dataset.csv")
```

## Dataset Review

This code is used to find the date range of the dataset.
```{r}
#Code to find the earliest and latest dates
data[["date"]] <- as.Date(data[["date"]])

# Find the earliest and latest dates
earliest_date <- min(data[["date"]], na.rm = TRUE)
latest_date <- max(data[["date"]], na.rm = TRUE)

# Print results
print(earliest_date)
print(latest_date)
```

**When was the data collected?** The data was collected in 2014. The time range is from 5/2/2014 to 7/10/2014. The data was uploaded to kaggle and last updated seven months ago.

**Where was the data acquired?** The data was acquired from kaggle.com. The data found on that website was originally on zillow.com, the real estate website.

**How was the data acquired?** Zillow’s Economic Research Team gathers, refines, and publishes housing and economic data from both public and proprietary sources. The core of Zillow’s data is derived from public property records filed with local municipalities, including deeds, property details, parcel information, and transaction histories. Many of the statistics in these datasets are calculated from raw property data. The methodology behind these calculations is further explained in the next section, where the dataset attributes are described.

**What are the attributes of this dataset?** The dataset consists of 18 attributes that describe various characteristics of properties. The Price variable represents the sale price of the property in USD and serves as the target variable. Property features such as Bedrooms, Bathrooms, Sqft Living, Sqft Lot, and Floors provide insight into the size and layout of the home. The View and Condition variables are ordinal indices that rate the property’s visual appeal and overall state. The Waterfront attribute is a nominal, binary indicator of whether the property has a waterfront view. Additionally, the dataset includes Yr Built and Yr Renovated, which record the year the property was originally constructed and last updated. Location-based attributes, including Street, City, Statezip, and Country, offer contextual information on where the property is situated.

**What type of data do these attributes contain?** The dataset contains a mix of nominal, ordinal, interval, and ratio data types. Nominal variables include categorical data without an inherent order, such as Street, City, Statezip, Country, and Waterfront. Ordinal variables represent ranked attributes with meaningful order but uneven intervals, such as View, Condition, and Date. Interval data, such as Yr Built and Yr Renovated, have meaningful differences between values but no true zero point. Finally, Ratio data includes numerical attributes with a true zero, such as Price, Bedrooms, Bathrooms, Sqft Living, Sqft Lot, Floors, Sqft Above, and Sqft Basement, allowing for meaningful comparisons and calculations.


## Data Dictionary

This code creates a table that shows the meaning of each variable included in the dataset.

```{r}
#import data dictionary
data_dictionary = read.csv("data_dictionary.csv")

library(knitr)
library(kableExtra)

# Create a nice table with kable and kableExtra
kable(data_dictionary, caption = "Variable Descriptions", col.names = c("Variable", "Description", "Type")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, italic = TRUE) %>%
  row_spec(0, background = "lightgray")  # Add a gray background to the header


```

## Summary of Statistics

This code creates a summary of the variables, including count, mean, standard deviation, min, Q1, Median, Q3, and max.

```{r}

# Load necessary libraries
library(dplyr)
library(readr)
library(gt)
library(tidyr)  # Needed for pivoting data

# Select numeric columns
numeric_data <- data %>%
  select(where(is.numeric))

# Generate summary statistics using reframe() to avoid errors
numeric_summary <- numeric_data %>%
  reframe(
    Variable = names(.),
    Count = sapply(., function(x) sum(!is.na(x))),
    Mean = sapply(., mean, na.rm = TRUE),
    SD = sapply(., sd, na.rm = TRUE),
    Min = sapply(., min, na.rm = TRUE),
    Q1 = sapply(., quantile, 0.25, na.rm = TRUE),
    Median = sapply(., median, na.rm = TRUE),
    Q3 = sapply(., quantile, 0.75, na.rm = TRUE),
    Max = sapply(., max, na.rm = TRUE)
  )

# Create a nicely formatted table with gt
numeric_summary %>%
  gt() %>%
  tab_header(title = "Summary Statistics - Numeric Variables") %>%
  fmt_number(columns = -1, decimals = 2) %>%
  tab_options(
    table.border.top.color = "black",
    table.border.bottom.color = "black",
    row.striping.include_table_body = TRUE
  )


```

# Updated Summary Statistics

This code updates the summary statistics with the mode and any missing data.

```{r}
# Load necessary libraries (some already loaded earlier, added here for completeness)
library(dplyr)
library(readr)
library(kableExtra)

# Function to calculate mode
calculate_mode <- function(x) {
  unique_x <- na.omit(x)
  if (length(unique_x) == 0) return(NA)
  mode_value <- unique_x[which.max(tabulate(match(x, unique_x)))]
  return(mode_value)
}

# Use the existing 'data' dataframe loaded from "USA Housing Dataset.csv"
df <- data

# Select numeric columns
numeric_df <- df %>%
  select(where(is.numeric))

# Select nominal (categorical) columns
# Note: Assuming 'street', 'city', 'statezip', 'country' are character columns as per dataset description
nominal_df <- df %>%
  select(where(is.character))

# Generate summary statistics for numeric variables
numeric_summary <- numeric_df %>%
  reframe(
    Variable = names(.),
    Count = sapply(., function(x) sum(!is.na(x))),
    Missing = sapply(., function(x) sum(is.na(x))),
    Mean = sapply(., mean, na.rm = TRUE),
    SD = sapply(., sd, na.rm = TRUE),
    Min = sapply(., min, na.rm = TRUE),
    Q1 = sapply(., quantile, 0.25, na.rm = TRUE),
    Median = sapply(., median, na.rm = TRUE),
    Q3 = sapply(., quantile, 0.75, na.rm = TRUE),
    Max = sapply(., max, na.rm = TRUE)
  )

# Generate summary statistics for nominal variables
nominal_summary <- nominal_df %>%
  reframe(
    Variable = names(.),
    Count = sapply(., function(x) sum(!is.na(x))),
    Missing = sapply(., function(x) sum(is.na(x))),
    Mode = sapply(., calculate_mode)
  )

# Create formatted tables with kableExtra
numeric_summary %>%
  kable(format = "html", digits = 2, caption = "Summary Statistics - Numeric Variables") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE)

nominal_summary %>%
  kable(format = "html", caption = "Summary Statistics - Nominal Variables") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE)
```

## Boxplot Comparing Waterfront & Non-Waterfront

This code creates a boxplot that will show the prices of houses that are on the water versus not on the water.

```{r}
library(ggplot2)

# Create the boxplot with outliers
ggplot(data, aes(x = as.factor(waterfront), y = price)) +
  geom_boxplot(fill = c("skyblue", "orange"), alpha = 0.7, outlier.color = "red", outlier.shape = 16) +
  labs(
    x = "Waterfront (0 = No, 1 = Yes)",
    y = "House Price",
    title = "House Prices: Waterfront vs. Non-Waterfront (With Outliers)"
  ) +
  theme_minimal()

```

**Analysis**
This visual is difficult to analyze because of the scale of the data. There are clearly some outliers included that, when removed, may make it easier to look for insight in the data.

## Outlier Analysis

This code is used to calculate outliers and average them for waterfron and non-waterfront houses.

```{r}
library(dplyr)

# Calculate IQR for non-waterfront houses (waterfront = 0)
non_waterfront_stats <- data %>%
  filter(waterfront == 0) %>%
  summarise(
    Q1 = quantile(price, 0.25, na.rm = TRUE),
    Q3 = quantile(price, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    Lower_Bound = Q1 - 1.5 * IQR,
    Upper_Bound = Q3 + 1.5 * IQR
  )

# Identify outliers and non-outliers
non_waterfront_outliers <- data %>%
  filter(waterfront == 0 & (price < non_waterfront_stats$Lower_Bound | price > non_waterfront_stats$Upper_Bound))

non_waterfront_non_outliers <- data %>%
  filter(waterfront == 0 & (price >= non_waterfront_stats$Lower_Bound & price <= non_waterfront_stats$Upper_Bound))

# Calculate the average sqft_lot
avg_sqft_lot_outliers <- mean(non_waterfront_outliers$sqft_lot, na.rm = TRUE)
avg_sqft_lot_non_outliers <- mean(non_waterfront_non_outliers$sqft_lot, na.rm = TRUE)

# Print results
cat("Average sqft_lot of non-waterfront outliers:", avg_sqft_lot_outliers, "\n")
cat("Average sqft_lot of non-waterfront non-outliers:", avg_sqft_lot_non_outliers, "\n")
```

This shows the massive difference between non-waterfront house outliers and non-outliers. The outliers have much more square footage on average than the non-waterfront houses.

## Updated Box Plot

This plot now showcases the data without the outliers
```{r}
library(ggplot2)
library(dplyr)

# Calculate IQR and filter out outliers
filtered_data <- data %>% 
  group_by(waterfront) %>% 
  mutate(
    Q1 = quantile(price, 0.25, na.rm = TRUE),
    Q3 = quantile(price, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    Lower_Bound = Q1 - 1.5 * IQR,
    Upper_Bound = Q3 + 1.5 * IQR
  ) %>%
  filter(price >= Lower_Bound & price <= Upper_Bound) %>%
  ungroup()

# Create the boxplot without outliers
ggplot(filtered_data, aes(x = as.factor(waterfront), y = price)) +
  geom_boxplot(fill = c("skyblue", "orange"), alpha = 0.7) +
  labs(
    x = "Waterfront (0 = No, 1 = Yes)",
    y = "House Price",
    title = "House Prices: Waterfront vs. Non-Waterfront (Without Outliers)"
  ) +
  theme_minimal()
```

### Boxplot Interpretation

The boxplot compares house prices between waterfront and non-waterfront properties:

1. **Price Distribution**:
   - Waterfront properties (1) have a significantly higher median house price compared to non-waterfront properties (0).
   - The interquartile range (IQR) for waterfront properties is larger, indicating more variation in prices.

2. **Outliers**:
   - Non-waterfront properties show some high-price outliers, but the overall price range is much lower.
   - Waterfront properties have a higher maximum price, even when outliers are removed.

3. **Conclusion**:
   - The data suggests that being located on the waterfront is a strong predictor of higher house prices.
   - The price disparity indicates that buyers may be willing to pay a premium for waterfront locations.

# Additional Data Source for Investment Strategy

## Real Estate Market Trends Dataset

One useful additional dataset for informing an investment strategy would be a real estate market trends dataset, such as **Zillow’s Housing Data** or the **Federal Housing Finance Agency (FHFA) House Price Index**.

### Why would this dataset be useful?

This dataset would provide insights into housing price trends over time, mortgage rates, and regional market fluctuations. Understanding historical trends can help predict future property values and assess whether a given area is appreciating or depreciating in value.

### How could it complement the data you are currently analyzing?

The current dataset focuses on specific property features (e.g., waterfront status) and their impact on prices. However, macroeconomic factors, such as interest rates, inflation, and supply-demand dynamics, also influence real estate prices. A market trends dataset would provide a broader economic context to better assess investment potential.

### Additional Dataset Link:
- [Zillow Research Data](https://www.zillow.com/research/data/)

